%!TEX root = ../calculus_tutorial.tex

	\subsection{Applications of integration}

		Integrals are essential tools used in many areas of science, engineering, and business.

		\subsubsection{Kinematics}

    			Calculus was originally invented
			to describe the equations of motion $x(t)$, $v(t)$, and $a(t)$,
			which correspond to an object's
			\emph{position}, \emph{velocity}, and \emph{acceleration} at time $t$.
			We call these the \emph{kinematics} equations,
			from the Greek word \emph{kinema} for motion.
			The velocity function $v(t)$ is the derivative of the position function,
			and the acceleration $a(t)$ is the derivative of the velocity.
			We can summarize these relations as follows:
			\[
				a(t) \overset{\frac{d}{dt} }{\longleftarrow} v(t) \overset{\frac{d}{dt} }{\longleftarrow} x(t).
			\]
			Newton's second law tells us that the acceleration of an object of mass $m$
			that has a net force $F_{\textrm{net}}$ acting on it is
			$a = \tfrac{1}{m}F_{\textrm{net}}$.
			Given the knowledge of acceleration over time $a(t)$,
			we can predict the position of the object $x(t)$ at any time $t$
			by ``undoing'' the two derivative operations using integration:
			\[
				% \underbrace{ 
				  \tfrac{1}{m} F_{\textrm{net}}(t)
				% }_{\text{dynamics}}
				\; = \; 
				\underbrace{ 
				  a(t) \; \overset{v_i + \int\!dt }{\longrightarrow} \; 
				  v(t) \; \overset{x_i + \int\!dt }{\longrightarrow} \; 
				  x(t) 
				}_{\text{kinematics}}.
			\]
			We integrate $a(t)$ to obtain $v(t)$
			and choose the initial velocity $v_i$ as the integration constant so that $v(0) = v_i$.
			We then use integration a second time to obtain $x(t)$ from $v(t)$,
			using the initial position $x_i = x(0)$ as the integration constant.			

			The case of \emph{uniform accelerated motion} (UAM) is of particular interest.
			% since it describes the trajectory of objects falling under the effect of gravity.
			%	Consider an object of mass $m$ that has a constant net $F_{\textrm{net}}$
			%	acting on it will experience 
			Consider an object that experiences a constant acceleration $a(t) = a$.
			We can use integration to find the velocity of this object at a later time $t=\tau$:
			\[ 
				v(\tau)	= v_i + \int_0^\tau a(t)\,dt
						= v_i +  \int_0^\tau a \,dt
						= v_i + a\tau.
			\]
			% where $v_i$ is the initial velocity of the object at $t=0$.
			Knowing the velocity as a function of time $v(t)$,
			we can use integration a second time to find its position at time $\tau$:
			\[
				x(\tau)	= x_i + \int_0^\tau v(t) \, dt
						= x_i +  \int_0^\tau (v_i+at) \, dt
						= x_i + v_i\tau + \tfrac{1}{2}a\tau^2.
			\]
			These two calculus steps
			allow us to obtain the famous kinematics equation
			$x(t) = x_i + v_it + \tfrac{1}{2}at^2$
			that describes the motion of objects undergoing constant acceleration $a$. % $a(t) = a$.
			Students taking a physics class are normally presented with this equation
			and it seems to come out of nowhere,
			but if you understand calculus,
			you'll know where it comes from:
			the integration operation applied twice to the constant acceleration function $a(t) = a$.
			% and the initial conditions $x_i \eqdef x(0)$ and $v_i \eqdef v(0)$.

			% don't have memorize it,
			% you can always derive from first principles $a(t) = a$,
			%
			%	Note that the above calculations required knowing the initial conditions $x_i$ and $v_i$.
			%	These initial values were required
			%	because the integral calculations we performed
			%	only tell us the \emph{change} in the quantities relative to their initial values.


		\subsubsection{Solving differential equations}

			Many important laws in science are described by \emph{differential equations}.
			A differential equation specifes an unknown function $f(t)$
			in terms of their derivatives $f^{\prime\!}(t)$, $f^{\prime\prime\!}(t)$, etc.
			Integrals are the primary tool we use to solve differential equations.
			% MAYBE diff equation = description of a system, and an instant later    — via Infinite Powers > Intro

			Here are some examples of differential equations used in science
			and their solutions:

			\begin{itemize}

				\item	The kinematics equations when the acceleration is constant
					come from the differential equation $x^{\prime\prime\!}(t) = a$.
					We use integration twice to find the unknown function
					$x(t) = x_i + v_it + \tfrac{1}{2}at^2$.
					We can verify that $x(t)$ is a solution to the differential equation $x^{\prime\prime\!}(t) = a$
					by computing the second derivative of $x(t)$.
					% which is $x^{\prime\prime\!}(t) = a$.

				\vspace{2mm}	% LAYOUT

				\item In biology,
					unconstrained bacterial growth is described by the equation $b^{\prime\!}(t) = kb(t)$,
					where $b(t)$ is the number of bacteria at time $t$.
					Intuitively,
					the bacterial growth rate $b^{\prime\!}(t)$
					is proportional to the number of existing bacteria.
					The solution to this equation is $b(t) = b_0e^{k t}$,
					where $b_0$ describes the number of bacteria at time $t = 0$.

				\vspace{2mm}	% LAYOUT

				\item Radioactive decay is described by the differential equation
					$r^{\prime\!}(t) = -\lambda r(t)$,
					where $r(t)$ describes the number of atoms of some radioactive element.
					The solution is $r(t) = r_0e^{-\lambda t}$.

				\vspace{2mm}	% LAYOUT

				\item Simple harmonic motion is described by the
					differential equation $x^{\prime\prime\!}(t) + \omega^2 x(t) = 0$,
					which has solution $x(t) = c_1 \sin(\omega t) + c_2 \cos(\omega t)$,
					for some constants $c_1$ and $c_2$.

			\end{itemize}

			\noindent
			If you take a course on differential equations,
			you'll learn all kind of tricks and techniques for solving differential equations.
			Integration plays a key role in all these techniques,
			since it allows us to ``undo'' the derivative operation.			
			%	We use integrals to solve differential equations.
			%	If we have to solve for $f(x)$ in the differential equation $\frac{d}{dx}f(x) = g(x)$,
			%	we can take the integral on both sides of the equation to obtain the answer $f(x) = \int g(x)\,dx + C$.

		\smallskip
		We also use integrals whenever we want to
		compute the total accumulation of some quantity that varies over time or space.
		For example,
		we can use integration to compute the total mass of an object
		whose mass density varies from point to point.


		\subsubsection{Probability and expectation calculations}

			Integration is required for computing probabilities of continuous random variables.
			A continuous random variable $X$
			is described by its \emph{probability density function}~$f_X$.
			To compute the probability of the event $\{ a \leq X \leq b\}$,
			we must calculate the total probability density $f_X$
			that falls between $x=a$ and $x=b$,
			which is given by the following integral:
			\[
				\Pr( \{ a \leq X \leq b\} )
					\;\; \eqdef \;\; \int_a^b f_X(x) \, dx.
			\]
			%	The probability density $f_X$ varies for different values of $x$,
			%	so if we want to compute the total probability of $X$ 
			%	we must compute the integral of $f_X$.
			%
			For example,
			the standard normal random variable $Z$
			is described by the probability density function
			$f_Z(z) = \tfrac{1}{\sqrt{2\pi}} e^{ -\frac{1}{2}z^2 }$.
			The probability of the event $\{ -1 \leq Z \leq 1\}$
			is calculated as the integral $\int_{-1}^1 f_Z(z) \, dz = \tt{quad(}f_Z\tt{,a=-1,b=1)[0]} = 0.68269$.
			% MAYBE: this means 68\% of outcomes of the random variable Z will be between -1 and 1


			\ifthenelse{\boolean{FORSTATSBOOK}}{
				The expected value of the quantity $G=g(X)$
				under the randomness of a continuous random variable $X$
				is defined as the following integral calculation:
				$\EE_X[g(X)] \eqdef \int_{x \in \calX} g(x)f_X(x)\,dx$.
				The expected value is computed by ``weighing'' each value of $g(x)$
				by the corresponding probability density for the event $\{X=x\}$,
				summed over all possible values for the random variable $X$.
				The mean $\mu = \EE_X[X]$ and the variance $\sigma^2 = \EE_X[(X-\mu)^2]$
				are two central concepts in probability theory and statistics
				that are computed as expectation integrals.
				Every time we use the $\EE_X$ notation in Section~\ref{sec:continuous_prob_distr},
				there is some integral calculation going on behind the scenes,
				so its good if you know to know a thing or two about integrals.
			}{
				We also use integration to compute \emph{expectations}.
				For example,
				the mean of the random variable $Z$
				is defined as the following integral calculation
				$\mu_Z \eqdef \EE[Z] = \int_{-\infty}^\infty zf_Z(z) \, dz$,
				and its variance is defined as 
				$\sigma_Z^2 \eqdef \EE[(Z-\mu_Z)^2] = \int_{-\infty}^\infty (Z-\mu_Z)^2 f_Z(z) \, dz$.
				The mean and the variance are central concepts in probability theory and statistics,
				so your integration skills will come in handy when learning these subjects.
			}


% OTHER IDEAS
%	* **Volume of strange shapes**
%	  * Build a 3D object out of thin slices.
%	  * Example: volume of a wine glass, vase, or trumpet bell.
%	
%	* **Centre of mass**
%	  * Integration locates the balance point of an object with uneven shape or density.
%	  * Example: where does an L-shaped metal plate balance on one finger?
%	
%	* **Work done by a variable force**
%	  * If force changes with position, $W=\int F(x),dx$.
%	  * Cool example: energy needed to compress a spring or pump water out of a tank.
%	
%	* **Fluid pressure and hydrostatics**
%	  * Pressure changes with depth, so total force on a dam or submarine hull comes from integration.
%	  * Example: why the bottom of a dam must be much thicker than the top.
%	
%	* **Electric charge, electric field, gravitational field**
%	  * Continuous distributions are handled by integrating small contributions.
%	  * Example: gravitational pull of a ring, disk, or planet on a nearby object.
%	
%	* **Signal processing**
%	  * Integrals measure accumulated signal strength or energy.
%	  * Example: total energy in an audio waveform or radio signal.
%	
%	* **Population growth and epidemiology**
%	  * Add up changing rates over time.
%	  * Example: total number of infections generated over a wave of an epidemic.
%	
%	* **Economics**
%	  * Recover totals from marginal quantities.
%	  * Example: total cost from marginal cost, or consumer surplus as area under a demand curve.
%	
%	* **Medicine and pharmacokinetics**
%	  * Drug concentration varies over time; integration gives total exposure.
%	  * Example: AUC = area under the concentration-time curve.
%	
%	* **Computer graphics**
%	  * Light hitting a surface can be modelled by integrating contributions from many directions.
%	  * Cool example: realistic shadows and soft lighting in rendered movies and games.
%	
%	* **Machine learning**
%	  * Integrals appear in continuous probability models and Bayesian inference.
%	  * Example: marginalizing over unknown parameters.
%	
%	* **Arc length and surface area**
%	  * Integration measures curved geometry.
%	  * Example: length of a roller-coaster track or area of a curved dome.
%	
%	* **Heat flow and diffusion**
%	  * Integrate changing heat or particle flux.
%	  * Example: how much total heat enters a material during warming.

